Precise Asymptotics of Information Transfer in Two-layer Linear Neural Networks

### 2021-06-30 revision_journal/results.tex
update intro wip
intro wip
sec 2 wip

\paragraph{Examples.}
To describe quantitively the bias-variance tradeoff phenomenon and the effect of varying sample sizes on the predication loss, we consider a \emph{random-effect model}, which has been studied for  single-task linear regression and ridge regression (see e.g. \cite{dobriban2020wonder,dobriban2018high}).

 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
\begin{example}%[Sample sizes]
\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
	Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, we let
$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, $$
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.

\begin{example} %[Covariate shift]
\label{ex_covshift}

%Our second example focuses on how varying covariate shifts impacts the \textit{variance} limit in equation \eqref{lem_cov_shift_eq}.
%Our first observation is that:
However, we now use an example to show that the above observation fails when $n_1$ is comparable to $n_2$. We take $\mu =1$, and consider the subset of covariate shift matrices satisfying $M=M^{-1}$. In other words, half of the singular values of $M$ satisfy that $\lambda_1 \ge \lambda_2\ge \cdots \ge \lambda_{p/2} \ge 1$, while the other half eigenvalues are $\lambda_{p/2}^{-1}\ge \cdots \ge \lambda_2^{-1} \ge \lambda_1^{-1}$. When $\lambda_1/\lambda_{p/2} = 1$, we have $M=\id_{p\times p}$ and there is no covariate shift.
%while as As $\lambda$ increases, the severity of covariate shift increases.
We claim the following dichotomy.
\begin{enumerate}
	\item If $n_1 \ge n_2$, then the variance limit is smallest when there is no covariate shift.
	\item If $n_1 < n_2$, then the variance limit is largest when there is no covariate shift.
\end{enumerate}
We explain why the above dichotomy happens. We can write the variance limit as
$$\ell_{\var}(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} a_1 + a_2} + \frac1{\lambda_i^{-2} a_1 + a_2} \right) .$$
When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
$$\ell_{\var}(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma} , $$
where we abbreviate $\gamma:=p/(n_1+n_2)$. Then using $a_1+a_2=1-\gamma$, through a direct calculation we find that %we can calculate that
\begin{align*}
\ell_{\var}(M) - \ell_{\var}(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
&= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 a_1(M) \left[ a_1(M) - a_2(M)\right] }{[a_1(M)+\lambda_i^2 a_2(M)][\lambda_i^2 a_1(M)+a_2(M)]} .
\end{align*}
%in this example is equal to
%$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
%\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
%Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
%\begin{align*}
%	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
%	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
%	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
%	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
%\end{align*}
%\begin{align*}
%	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
%\end{align*}
%where $g(\lambda, a_1) \ge 0$.
%and can be derived from algebraic calculations (details omitted).
We claim that $a_1(M) > a_2 (M)$ if and only if $n_1 > n_2$, which then explains the dichonomy. In fact, if $a_1>a_2$, then the first equation in \eqref{eq_a12extra} gives that $a_1> (1-\gamma)/2$, and the second equation in  \eqref{eq_a12extra} gives that
\begin{align*}
 \frac{n_1}{n_1 + n_2} &> a_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2}.
\end{align*}
%\begin{align*}
% \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
%	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
%\end{align*}
This implies $n_1>n_2$. The other direction follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
\end{example}

\subsection{Task size imbalance}\label{sec_sizeratio}

@@

Next, we use the bias-variance limits to study how varying sample sizes impacts HPS.
For example, imagine if we want to decide whether to collect more of task one's data or not, how does increasing $n_1$ affect the prediction loss?
We assume that $n_2$ is fixed for simplicity.
%Now we illustrate an interesting phenomenon that adding task one's samples helps task two initially, but may hurt eventually.
The variance limit in equation \eqref{cor_MTL_error} obvious decreases with $n_1$.
It turns out that the bias term always increases with $n_1$, which can be verified by showing that the bias limit's derivative is always nonnegative.
%As a function of the sample ratio, the limiting estimate always decreases first from $\frac{\sigma^2 p}{n_2 - p}$ with $n_1$ being zero, and then increases to $d^2$ when $n_1$ goes to infinity.
%We describe a sketch of the proof.
By comparing the derivative of the bias and variance limits with respect to $n_1$ (details omitted), we obtain the following dichotomy.
\begin{enumerate}
	\item When $\frac{d^2}{\sigma^2} < \frac{p}{4n_2 - 6p}$, the prediction loss decreases monotonically as $n_1$ increases.
	Intuitively, this regime of $d^2$ always helps task two.
	\item When $\frac{d^2}{\sigma^2} > \frac{p}{4n_2 - 6p}$, the prediction loss always decreases first from $\frac{\sigma^2 p}{n_2 - p}$ (when $n_1 = 0$), and then increases to $d^2$ (when $n_1 \rightarrow \infty$).
	To see this, near the point where $n_1$ is zero, one can verify (from the derivatives) that bias increases less while variance decreases more, and there is \textit{exactly} one critical point where the derivative is zero, which corresponds to the \textit{optimal sample size ratio}.
\end{enumerate}
\fi
	\end{example}

@@

>>>>

\subsection{Uneven sample sizes}\label{sec_sizeratio}

@@

Next, we use the bias-variance limits to study how varying sample sizes impacts HPS.
For example, imagine if we want to decide whether to collect more of task one's data or not, how does increasing $n_1$ affect the prediction loss?
We assume that $n_2$ is fixed for simplicity.
%Now we illustrate an interesting phenomenon that adding task one's samples helps task two initially, but may hurt eventually.
The variance limit in equation \eqref{cor_MTL_error} obvious decreases with $n_1$.
It turns out that the bias term always increases with $n_1$, which can be verified by showing that the bias limit's derivative is always nonnegative.
%As a function of the sample ratio, the limiting estimate always decreases first from $\frac{\sigma^2 p}{n_2 - p}$ with $n_1$ being zero, and then increases to $d^2$ when $n_1$ goes to infinity.
%We describe a sketch of the proof.
By comparing the derivative of the bias and variance limits with respect to $n_1$ (details omitted), we obtain the following dichotomy.
\begin{enumerate}
	\item When $\frac{d^2}{\sigma^2} < \frac{p}{4n_2 - 6p}$, the prediction loss decreases monotonically as $n_1$ increases.
	Intuitively, this regime of $d^2$ always helps task two.
	\item When $\frac{d^2}{\sigma^2} > \frac{p}{4n_2 - 6p}$, the prediction loss always decreases first from $\frac{\sigma^2 p}{n_2 - p}$ (when $n_1 = 0$), and then increases to $d^2$ (when $n_1 \rightarrow \infty$).
	To see this, near the point where $n_1$ is zero, one can verify (from the derivatives) that bias increases less while variance decreases more, and there is \textit{exactly} one critical point where the derivative is zero, which corresponds to the \textit{optimal sample size ratio}.
\end{enumerate}
\fi
%	\end{example}

@@

\todo{describe comparison to the TL estimator and the weighted OLS estimator.}

\paragraph{Examples.}
To describe quantitatively the bias-variance tradeoff phenomenon and the effect of varying sample sizes on the predication loss, we consider a \emph{random-effect model}, which has been studied for  single-task linear regression and ridge regression (see e.g. \cite{dobriban2020wonder,dobriban2018high}).

 %{How does hard parameter sharing scale with sample sizes and covariate shift $M$?} One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift. Next, we illustrate how varying them impact the prediction loss.
%\begin{example}%[Sample sizes]
%\label{ex_sample_ratio}
	%We first consider the impact of varying sample sizes.
	%Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.
	%We again consider the random effect model in Example \ref{ex_same_cov} with $t=2$.
	Suppose each $\beta^{(i)}$ consists of two random components, one that is shared among the two tasks and one that is task-specific. More precisely, we let
$$\beta^{(i)}=\beta_0 +\wt \beta^{(i)},\quad i=1,2, $$
where $\beta_0$ denotes the shared component, %whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1}\kappa^2$,
and $\wt \beta^{(i)}$ denotes the $i$-th
%Let $\beta^{(i)}$ be equal to $\beta_0$ plus a
task-specific component whose entries are i.i.d. Gaussian random variables of mean zero and variance $p^{-1} d^2$.

%\begin{example} %[Covariate shift]
%\label{ex_covshift}

%Our second example focuses on how varying covariate shifts impacts the \textit{variance} limit in equation \eqref{lem_cov_shift_eq}.
%Our first observation is that:
However, we now use an example to show that the above observation fails when $n_1$ is comparable to $n_2$. We take $\mu =1$, and consider the subset of covariate shift matrices satisfying $M=M^{-1}$. In other words, half of the singular values of $M$ satisfy that $\lambda_1 \ge \lambda_2\ge \cdots \ge \lambda_{p/2} \ge 1$, while the other half eigenvalues are $\lambda_{p/2}^{-1}\ge \cdots \ge \lambda_2^{-1} \ge \lambda_1^{-1}$. When $\lambda_1/\lambda_{p/2} = 1$, we have $M=\id_{p\times p}$ and there is no covariate shift.
%while as As $\lambda$ increases, the severity of covariate shift increases.
We claim the following dichotomy.
\begin{enumerate}
	\item If $n_1 \ge n_2$, then the variance limit is smallest when there is no covariate shift.
	\item If $n_1 < n_2$, then the variance limit is largest when there is no covariate shift.
\end{enumerate}
We explain why the above dichotomy happens. We can write the variance limit as
$$\ell_{\var}(M)=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2}\left( \frac{1}{\lambda_i^{2} a_1 + a_2} + \frac1{\lambda_i^{-2} a_1 + a_2} \right) .$$
When $M=\id_{p\times p}$, by the first equation of \eqref{eq_a12extra}, we have
$$\ell_{\var}(\id_{p\times p})=\frac{\sigma^2}{ n_1+n_2 }\sum_{i=1}^{p/2} \frac{2}{1-\gamma} , $$
where we abbreviate $\gamma:=p/(n_1+n_2)$. Then using $a_1+a_2=1-\gamma$, through a direct calculation we find that %we can calculate that
\begin{align*}
\ell_{\var}(M) - \ell_{\var}(\id_{p\times p})%&= \frac{\sigma^2 \gamma}{2(1-\gamma)} (\lambda^2-1)a_1\cdot \bigbrace{  \frac{1}{ -a_1(\lambda^2-1)+(1-\gamma)\lambda^2 } - \frac{1}{a_1(\lambda^2-1) + (1-\gamma)}} \\
&= \frac{\sigma^2 }{ n_1+n_2-p} \sum_{i=1}^{p/2} \frac{(\lambda_i^2-1)^2 a_1(M) \left[ a_1(M) - a_2(M)\right] }{[a_1(M)+\lambda_i^2 a_2(M)][\lambda_i^2 a_1(M)+a_2(M)]} .
\end{align*}
%in this example is equal to
%$\frac{p}{2(n_1 + n_2)} f(\lambda)$, where
%\[ f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}. \]
%Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2}$, we can verify
%\begin{align*}
%	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
%	&+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1} \\
%	&- \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
%	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
%\end{align*}
%\begin{align*}
%	f(\lambda) - f(1) &= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)  g(\lambda, a_1), %\cdot (\lambda^2-1)^2
%\end{align*}
%where $g(\lambda, a_1) \ge 0$.
%and can be derived from algebraic calculations (details omitted).
We claim that $a_1(M) > a_2 (M)$ if and only if $n_1 > n_2$, which then explains the dichonomy. In fact, if $a_1>a_2$, then the first equation in \eqref{eq_a12extra} gives that $a_1> (1-\gamma)/2$, and the second equation in  \eqref{eq_a12extra} gives that
\begin{align*}
 \frac{n_1}{n_1 + n_2} &> a_1 + \frac{1}{n_1+n_2} \sum_{i=1}^{p/2}\left(\frac{\lambda_i^2}{\lambda_i^2+1}+\frac{\lambda_i^{-2}}{\lambda_i^{-2}+1}\right) = \frac{1-\gamma}{2}+\frac{\gamma}{2}=\frac{1}{2}.
\end{align*}
%\begin{align*}
% \frac{n_1}{n_1 + n_2} &=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} \\
%	&> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) =\frac{1}{2}.
%\end{align*}
This implies $n_1>n_2$. The other direction follows from a similar argument. % Similarly, if $a_1<a_2$, equations  \eqref{eq_a12extra000} and  \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.
%\end{example}

\todo{draw a connection to \citet{david2010impossibility}}

\input{multitask.tex}

### 2021-06-30 revision_journal/setup.tex
sec 2 wip

Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$. For $i=1,2,$ Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
In this paper, we make the following assumptions on feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$. We remark that all these assumptions are natural in high-dimensional statistics.

@@

Another commonly used transfer learning estimator is defined as the minimizer of the total loss function \cite{bastani2020predicting, li2020transfer}:
$$\hat \beta^{\rm{TL}} := \argmin\left[ \norm{X^{(1)} \beta - Y^{(1)}}^2 + \norm{X^{(2)} \beta - Y^{(2)}}^2\right].$$

@@

%\FY{Introduce some popular transfer learning estimators: total loss estimator, and model averaging estimator.}

>>>>

Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$.
For $i=1,2,$ let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
Throughout the paper, we make the following assumptions on the feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$, following the work of \citet{hastie2019surprises} (see also \citet{wu2020optimal,richards2021asymptotics} and the references therein). %We remark that all these assumptions are natural in high-dimensional statistics.

@@

@@

%\FY{Introduce some popular transfer learning estimators: total loss estimator, and model averaging estimator.}
Another commonly used transfer learning estimator is defined as the minimizer of the total loss function \cite{bastani2020predicting, li2020transfer}:
$$\hat \beta^{\rm{TL}} := \argmin\left[ \norm{X^{(1)} \beta - Y^{(1)}}^2 + \norm{X^{(2)} \beta - Y^{(2)}}^2\right].$$
